---
title: "README"
author: "Imad-Alh"
date: "2024-09-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting and Cleaning Data project
This project is based on the Human Activity Recognition Using Smartphones dataset,of an ititial project by:
Jorge L. Reyes-Ortiz, Davide Anguita, Alessandro Ghio, Luca Oneto.

### Dataset Information

Jorge L. Reyes-Ortiz, Davide Anguita, Alessandro Ghio, Luca Oneto

Smartlab - Non Linear Complex Systems Laboratory

DITEN - Università degli Studi di Genova.

Genoa, Italy.

Email: activityrecognition@smartlab.ws

Website: www.smartlab.ws

## Objective

The objective of this project is to:

1- Merge the training and test sets into a single dataset.

2- Extract measurements on the mean and standard deviation for each observation.

3- Assign descriptive activity names to label the activities.

4- Label the dataset with appropriately descriptive variable names.

5- Create a tidy dataset with the average of each variable for each activity and each subject.

## Code Explanation

### 1. Reading and Merging the Data

The script begins by reading in the features and activity labels from the provided files. It then reads the training and test datasets (X_train, Y_train, X_test, Y_test) along with the corresponding subject IDs for both datasets. The training and test datasets are merged into one complete dataset using rbind().

```{r}
features <- read.table('UCI_HAR_Dataset\\features.txt')

X_train <- read.table('UCI_HAR_Dataset\\train\\X_train.txt')

colnames(X_train) <- features[,2]

Y_train <- read.table('UCI_HAR_Dataset\\train\\Y_train.txt')
colnames(Y_train) <- 'Labels' # setting the correct column name

X_train <- cbind(Y_train, X_train)

subjects_train <- read.table('UCI_HAR_Dataset\\train\\subject_train.txt')
colnames(subjects_train) <- 'Subject_ID'
X_train <- cbind(subjects_train, X_train)

X_test <- read.table('UCI_HAR_Dataset\\test\\X_test.txt')

colnames(X_test) <- features[,2]

Y_test <- read.table('UCI_HAR_Dataset\\test\\Y_test.txt')
colnames(Y_test) <- 'Labels' # setting the correct column name

X_test <- cbind(Y_test, X_test)

subjects_test <- read.table('UCI_HAR_Dataset\\test\\subject_test.txt')
colnames(subjects_test) <- 'Subject_ID'
X_test <- cbind(subjects_test, X_test)

complet_DF <- rbind.data.frame(X_train, X_test)
```

### 2. Extracting Mean and Standard Deviation Measurements

After merging the datasets, the script selects only those columns that represent mean and standard deviation measurements using the grep() function, which matches patterns in column names.

```{r}
mean_std_indices <- grep("mean\\(\\)|std\\(\\)", colnames(complet_DF))

complet_DF <- complet_DF[, c(1, 2, mean_std_indices)]
```

### 3. Labeling the Activities

Activity labels, which are provided as numeric codes, are replaced with descriptive activity names (WALKING, SITTING, etc.) using a mapping vector.

```{r}
activity_labels <- read.table('UCI_HAR_Dataset\\activity_labels.txt')
label_maping <- (activity_labels[,2])
names(label_maping) <- activity_labels[,1]

complet_DF$Labels <- label_maping[complet_DF$Labels]
```

### 4. Renaming Variables with Descriptive Names

The script renames the dataset columns with more descriptive and human-readable names. For example:

tBodyAcc-mean()-X becomes TimeBodyAccelerometerMean-X

fBodyGyro-std()-Z becomes FrequencyBodyGyroscopeStd-Z

```{r}
new_names <- c(
  "Subject_ID", "Activity_Label", 
  "TimeBodyAccelerometerMean-X", "TimeBodyAccelerometerMean-Y", "TimeBodyAccelerometerMean-Z", 
  "TimeBodyAccelerometerStd-X", "TimeBodyAccelerometerStd-Y", "TimeBodyAccelerometerStd-Z", 
  "TimeGravityAccelerometerMean-X", "TimeGravityAccelerometerMean-Y", "TimeGravityAccelerometerMean-Z", 
  "TimeGravityAccelerometerStd-X", "TimeGravityAccelerometerStd-Y", "TimeGravityAccelerometerStd-Z", 
  "TimeBodyAccelerometerJerkMean-X", "TimeBodyAccelerometerJerkMean-Y", "TimeBodyAccelerometerJerkMean-Z", 
  "TimeBodyAccelerometerJerkStd-X", "TimeBodyAccelerometerJerkStd-Y", "TimeBodyAccelerometerJerkStd-Z", 
  "TimeBodyGyroscopeMean-X", "TimeBodyGyroscopeMean-Y", "TimeBodyGyroscopeMean-Z", 
  "TimeBodyGyroscopeStd-X", "TimeBodyGyroscopeStd-Y", "TimeBodyGyroscopeStd-Z", 
  "TimeBodyGyroscopeJerkMean-X", "TimeBodyGyroscopeJerkMean-Y", "TimeBodyGyroscopeJerkMean-Z", 
  "TimeBodyGyroscopeJerkStd-X", "TimeBodyGyroscopeJerkStd-Y", "TimeBodyGyroscopeJerkStd-Z", 
  "TimeBodyAccelerometerMagnitudeMean", "TimeBodyAccelerometerMagnitudeStd", 
  "TimeGravityAccelerometerMagnitudeMean", "TimeGravityAccelerometerMagnitudeStd", 
  "TimeBodyAccelerometerJerkMagnitudeMean", "TimeBodyAccelerometerJerkMagnitudeStd", 
  "TimeBodyGyroscopeMagnitudeMean", "TimeBodyGyroscopeMagnitudeStd", 
  "TimeBodyGyroscopeJerkMagnitudeMean", "TimeBodyGyroscopeJerkMagnitudeStd", 
  "FrequencyBodyAccelerometerMean-X", "FrequencyBodyAccelerometerMean-Y", "FrequencyBodyAccelerometerMean-Z", 
  "FrequencyBodyAccelerometerStd-X", "FrequencyBodyAccelerometerStd-Y", "FrequencyBodyAccelerometerStd-Z", 
  "FrequencyBodyAccelerometerJerkMean-X", "FrequencyBodyAccelerometerJerkMean-Y", "FrequencyBodyAccelerometerJerkMean-Z", 
  "FrequencyBodyAccelerometerJerkStd-X", "FrequencyBodyAccelerometerJerkStd-Y", "FrequencyBodyAccelerometerJerkStd-Z", 
  "FrequencyBodyGyroscopeMean-X", "FrequencyBodyGyroscopeMean-Y", "FrequencyBodyGyroscopeMean-Z", 
  "FrequencyBodyGyroscopeStd-X", "FrequencyBodyGyroscopeStd-Y", "FrequencyBodyGyroscopeStd-Z", 
  "FrequencyBodyAccelerometerMagnitudeMean", "FrequencyBodyAccelerometerMagnitudeStd", 
  "FrequencyBodyBodyAccelerometerJerkMagnitudeMean", "FrequencyBodyBodyAccelerometerJerkMagnitudeStd", 
  "FrequencyBodyBodyGyroscopeMagnitudeMean", "FrequencyBodyBodyGyroscopeMagnitudeStd", 
  "FrequencyBodyBodyGyroscopeJerkMagnitudeMean", "FrequencyBodyBodyGyroscopeJerkMagnitudeStd"
)

colnames(complet_DF) <- new_names
```

### 5. Creating a Tidy Dataset

The final step is to group the data by Activity_Label and Subject_ID and calculate the average of each variable for each activity and each subject. This is done using the dplyr package's group_by() and summarise(across()) functions. The result is a tidy dataset that meets the project requirements.

```{r message=FALSE, warning=FALSE}
library(dplyr)
summary_df <- complet_DF %>%
  group_by(Activity_Label, Subject_ID) %>%
  summarise(across(everything(), \(x) mean(x, na.rm = TRUE)), .groups = "drop")
```

### 6. Saving the Output

The tidy dataset is saved as tidy_dataset.txt for further use.

```{r}
write.table(summary_df, file = "tidy_dataset.txt", row.names = FALSE)
```

## License:

Use of this dataset in publications must be acknowledged by referencing the following publication 

Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine. International Workshop of Ambient Assisted Living (IWAAL 2012). Vitoria-Gasteiz, Spain. Dec 2012

This dataset is distributed AS-IS and no responsibility implied or explicit can be addressed to the authors or their institutions for its use or misuse. Any commercial use is prohibited.

Jorge L. Reyes-Ortiz, Alessandro Ghio, Luca Oneto, Davide Anguita. November 2012.
